Secure Composition of Robust and Optimising Compilers

### 2025-02-05 icfp/sections/intro.tex
Rewrite intro

Compilers may fail to preserve these enforcements, even when proven correct or thoroughly tested~\cite{}.
While properties can be enforced at source-level by means of static analyses, after compiling to languages without such abstractions, the properties may break.
For example, RISC-V, a possible target-language of the most commonly used Rust compiler, is untyped and provides neither memory safety nor type safety guarantees.
Intuitively, this is why we want compilers to be {\em correct}, so that they provably preserve the meaning of the source-language.
Unfortunately, software is seldolmy developed in isolation and programs may be linked with target-level libraries after compilation.
These libraries can be compiled with different compilers or may even be hand-written, so the correctness-guarantees provided by one compiler may not hold.

This is why secure compilers have become an interesting avenue for research, but they are difficult to prove.

Compositionality to the rescue, proofs should compose to make it easier.

% a secure compiler, by definition, is supports source-independent linking
% vertical composition is a consequence of existence of backtranslations

- Demonstrate compositionality of secure compilation.

- Mechanization of core theory.

- Case Study demonstrating composition.

>>>>

Compilers may fail to preserve these enforcements, even when proven correct or thoroughly tested~\cite{}.
While properties can be enforced at source-level by means of static analyses, after compiling to languages without such abstractions, the properties may break.
For example, RISC-V, a possible target-language of the most commonly used Rust compiler,  provides neither memory nor type safety guarantees.
Intuitively, this is why we want compilers to be {\em correct}, so that they provably preserve the meaning of the source-language.
With preservation of meaning follows preservation of safety guarantees.
Unfortunately, software is seldolmy developed in isolation and programs may be linked with target-level libraries after compilation.
These libraries can be compiled with different compilers or may even be hand-written, so the correctness-guarantees provided by one compiler may not hold.

In previous work, it has been argued that correct compilation is a spectrum~\cite{} with different restrictions applying to how the compiled partial programs can interact with target-level libraries.
At the extreme end of the spectrum with basically zero restrictions\footnote{The minimal restriction that has to be applied is interface compatibility.} lies robust preservation of meaning.
Here, partial program components can interact with arbitrary contexts, i.e., libraries, and the compiler has to ensure that crossing the boundary between components and contexts does not break any properties whatsoever.
For example, in order to robustly ensure memory safety in Rust when compiling to RISC-V, the compiler could compartmentalize all calls and returns via the foreign function interface.
This compartmentalization can be done in different ways, one of which is via CHERI capabilities~\cite{}.

For the context of this paper, we are interested in formally verified compilers.
That is, they need to be proven to be robustly preserving, which is a labor-intensive and in parts difficult task~\cite{}. %lots of cits
Worse, compilers are not just simple functions translating from one domain into another. 
A practical compiler consists of several different compilation passes, which itself can be seen as compilers that translate from one intermediate representation into another.
The CompCert project~\cite{}, while having more restrictions on target-level contexts, attests the herculian effort necessary to prove correctness for realistic compilers with several different passes.
Moreover, for optimising passes, it can be beneficial to swap the order or iterate them until a fixpoint is reached, but this necessitates modifications to a top-level meaning-preservation proof of a compiler.
Therefore, it is helpful to have local meaning-preservation proofs for each pass and be able to compose them to obtain a whole-proof for the complete pipeline.

Compositionality of meaning-preservation proofs for compilers has been investigated in the context of different flavors of compiler correctness. 
But, to the best of our knowledge, compositionality for robustly preserving compilers is still unclear.
Moreover, there are two key compositionality properties~\cite{} that researchers care about: (1) vertical compositionality and (2) source-independent linking. 
Vertical compositionality asserts that it should be possible to verify a compiler pass in isolation.
For source-independent linking, note that when verifying a compiler, source and target terms need to be related somehow.
However, this is only necessary for the component, i.e., the partial program that is being compiled, and so source-independent linking asserts that the target context need not be related to any source-context.

Our aim with this paper is to lay the groundwork for secure compiler composition:
\begin{itemize}
  \item 
    We present a set of theorems (\Cref{sec:sequential}) that can be used for practical secure compiler verification.
    Intuitively, we prove that two compilers that robustly preserve two (possibly distinct) properties compose into a compiler that robustly preserves the intersection of the properties.
    We show special-cases, such as swapping secure optimisation passes or composition of compilers robustly preserving different properties. 
    Moreover, we discuss how robust preservation relates to vertical compositionality and source-independent linking.
  \item
    There is a mechanization of our core-theory in the Coq\footnote{To be renamed into Rocq.} proof assistant and we mark these instances with (\CoqSymbol) throughout the paper.
  \item 
    We provide an extensive case-study of different security properties (\Cref{sec:compprop}), a set of programming languages (\Cref{sec:casestud:defs}), and compilers between them (\Cref{sec:casestud:rtp}).
    The case-study consists of different pen-and-paper verified compiler passes comprised of robust preservation or enforcement mechanisms which individually ensure relevant security properties, such as memory safety in a speculative execution model.
\end{itemize}
The paper also provides background (\Cref{sec:background}), some formal insights (\Cref{sec:formalities}) and a discussion of related work (\Cref{sec:relwork}) before concluding (\Cref{sec:concl}).

Pen-and-paper proofs as well as the mechanization are available as supplementary material.
